sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-02 doc/paper-jss/inla-comparison.Rmd
Streamline chunk settings

```{r setup-inla-comparison, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  cache = TRUE,
  autodep = TRUE,
  warning = FALSE,
  message = FALSE
)
```

@@

```{r inla-pred, out.width="5.5in"}
idx <- inla.stack.index(sdat, "stdata")$data
d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]

ggplot(d) +
  geom_point(aes(X, Y, colour = inla_pred), size = 2) +
  facet_wrap(~year) +
  scale_colour_viridis_c() +
  coord_fixed()
```

>>>>

@@

```{r inla-pred, out.width="5.5in"}
# idx <- inla.stack.index(sdat, "stdata")$data
# d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]
# 
# ggplot(d) +
#   geom_point(aes(X, Y, colour = inla_pred), size = 2) +
#   facet_wrap(~year) +
#   scale_colour_viridis_c() +
  # coord_fixed()
```

### 2022-11-02 doc/paper-jss/paper-jss.Rmd
Streamline chunk settings

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  prompt = FALSE
  # R.options = list(prompt = " ", continue = " ")
)
```

@@

\pkg{sdmTMB} models can include penalized likelihoods by assigning priors to model parameters (`?\pkg{sdmTMB}priors`).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \pkg{TMB} [@breivik2021; @osgood-zimmerman2021], penalized complexity (PC) priors [@simpson2017; @fuglstad2019] (`?pc_matern`) can constrain the spatial range and variance parameters.

The results from an \pkg{sdmTMB} model can be used to generate various quantities of interest.
For example, \pkg{sdmTMB} provides functionality for generating population-level summaries for each time slice including the center of gravity  (`get_cog()`) [e.g., @thorson2016cog] and total population index (`get_index()`) given a user-supplied grid.

@@

```{r sdmTMB-lib0, eval=TRUE, echo=FALSE}
library(sdmTMB)
# simplify for paper:
pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |> 
  tibble::as_tibble()
```

@@

```{r pcod-eg1-suppress-depth2, eval=TRUE, echo=FALSE}
qcs_grid$depth_scaled <- NULL # for simplicity in paper
qcs_grid$depth_scaled2 <- NULL # for simplicity in paper
```

@@

@@

```{r owl-fit, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, eval=FALSE}
mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
fit_owls <- sdmTMB(
  count ~ nao + (1 | year_factor),
  spatial_varying = ~ 0 + nao,
  family = nbinom2(link = "log"), 
  data = snow, 
  mesh = mesh,
  time = "year", 
  spatial = "on", 
  spatiotemporal = "iid"
)
```

@@

Spatially and spatiotemporally explicit data are increasingly collected in ecology and have the power to reveal new ecological processes [e.g., @dinnage2020; @english2022] and improve ecological management [@sofaer2019].
These data present statistical challenges to modelling them effectively and efficiently since appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate.
We hope the development of user-friendly interfaces such as \pkg{sdmTMB} opens this useful class of models to a wider audience of users.

>>>>

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  prompt = FALSE,
  cache = TRUE,
  autodep = TRUE,
  fig.width = 7,
  fig.asp = 0.618,
  cache.comments = FALSE,
  dev = "png",
  dpi = 180
  # optipng = "-strip all"
  # R.options = list(prompt = "R> ", continue = "> ")
)
# knitr::knit_hooks$set(optipng = knitr::hook_optipng)
```

@@

\pkg{sdmTMB} models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in \pkg{TMB} [@breivik2021; @osgood-zimmerman2021], penalized complexity (PC) priors [@simpson2017; @fuglstad2019] (`?pc_matern`) can constrain the spatial range and variance parameters.

The results from an \pkg{sdmTMB} model can be used to generate various quantities of interest.
For example, for the field of ecology, \pkg{sdmTMB} provides functionality for generating population-level summaries for each time slice including the center of gravity  (`get_cog()`) [e.g., @thorson2016cog] and total population index (`get_index()`) given a user-supplied grid.

@@

```{r sdmTMB-lib0, eval=TRUE, echo=FALSE}
library(sdmTMB)
# simplify for paper:
pcod_original <- pcod # save it for later
pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |> 
  tibble::as_tibble()
```

@@

```{r pcod-eg1-suppress-depth2, eval=TRUE, echo=FALSE}
qcs_grid_original <- qcs_grid
qcs_grid$depth_scaled <- NULL # for simplicity in paper
qcs_grid$depth_scaled2 <- NULL # for simplicity in paper
```

@@

```{r pcod-restore, echo=FALSE}
pcod <- pcod_original
qcs_grid <- qcs_grid_original
```

@@

```{r owl-fit, warning=FALSE, message=FALSE, cache=TRUE, echo=TRUE, eval=FALSE}
mesh <- make_mesh(snow, xy_cols = c("X", "Y"), cutoff = 1.5)
fit_owls <- sdmTMB(
  count ~ nao + (1 | year_factor),
  spatial_varying = ~ nao,
  family = nbinom2(link = "log"), 
  data = snow, 
  mesh = mesh,
  time = "year", 
  spatial = "on", 
  spatiotemporal = "iid"
)
```

@@

<!-- Spatially and spatiotemporally explicit data are increasingly collected and have the power to reveal new ecological processes [e.g., @dinnage2020; @english2022] and improve ecological management [@sofaer2019]. -->
<!-- These data present statistical challenges to modelling them effectively and efficiently since appropriate models such as GLMMs with random fields are often computationally intensive and challenging to implement, interpret, and evaluate. -->
<!-- We hope the development of user-friendly interfaces such as \pkg{sdmTMB} opens this useful class of models to a wider audience of users. -->

### 2022-11-02 doc/paper-jss/pcod.Rmd
Ref apps
Streamline chunk settings

# Example: spatiotemporal modelling of Pacific cod

```{r setup-pcod, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  cache = TRUE,
  autodep = TRUE,
  cache.comments = FALSE,
  echo = TRUE
)
```

>>>>

# Example: spatiotemporal modelling of Pacific cod {#app:pcod}

```{r setup-pcod, include = FALSE, cache=FALSE}
knitr::opts_chunk$set(
  echo = TRUE
)
```
